Ionization of biological molecules by multicharged ions using the stoichiometric model

### 2019-11-20 ionmol_stoi.tex
response to reviewers and corrections

The ionization of biological molecules by multicharged ions constitutes 
the primary damage mechanism. The most widely used method to predict
such processes is the first Born approximation. At high energies, this 
perturbative method warrants the $Z^{2}$ laws, where $Z$ is the 
projectile charge. However, the damage is concentrated in the 
vicinities of the Bragg peak --at energies of hundreds of keV/amu--, 
precisely where the Born approximation starts to fail. 
Another theoretical issue arises due to the targets themselves; we are 
dealing with complex molecules, and the description of such targets 
represents a hard task for {\it ab initio} calculations. The objective 
of this article is to deal with these two aspects; first, we perform 
more appropriate calculations on the primary damage mechanism, which can 
replace the Born results. Second, we inspect and test a stoichiometric 
model to describe the ionization of molecular targets.

@@

The total ionization cross sections of these atoms were calculated using 
the CDW. The initial bound and final continuum radial wave 
functions were obtained by using the {\sc radialf} code, developed by 
Salvat and co--workers~\cite{salvat1995}, and a Hartree--Fock potential 
obtained from the Depurated Inversion Method~\cite{mendez2016,mendez2018}. 
We used a few thousand pivot points to solve the Schr\"{o}dinger 
equation, depending on the number of oscillations of the continuum 
state. The radial integration was performed using the cubic spline 
technique. We expand our final continuum wave function as usual,
\begin{equation}
\psi_{\overrightarrow{k}}^{-}(\overrightarrow{r})=\sum_{l=0}^{l_{\max
}}\sum_{m=-l}^{l}R_{kl}^{-}(r)Y_{l}^{m}(\widehat{r})Y_{l}^{m^{\ast }}
(\widehat{k})\,.
\label{eq:contwave}
\end{equation}
The number of angular momenta considered varied from 8, at 
very low ejected--electron energies, up to $l_{\max}\sim 30$, 
for the highest energies 
considered. The same number of azimuth angles were required to obtain 
the four--fold differential cross section. The calculation performed does 
not display prior--post discrepancies at all. Each atomic total cross 
section was calculated using 35 to 100 momentum transfer values, 28 
fixed electron angles, and around 45 electron energies depending on the 
projectile impact energy. 
Further details of the calculation are given in Ref.~\cite{montanari2017}.

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/atomicscaling.eps}
\caption{Reduced CDW total ionization cross section of six atomic 
targets. The curves are labeled with the charge state corresponding to 
the six multicharged projectiles.}
\label{fig:atomscaling}
\end{figure*}

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/ener_mean.eps}
\caption{Mean emitted energy distribution for ionization by the impact 
of multicharged ions, given by Eq.~(\ref{40}). 
Solid lines for ion charges $+1$, $2+$, $4+$, $6+$ and $8+$, as indicated.
Dashed lines for $\bar{p}$ and dotted line for Born with $Z=1$.}
\label{fig:emittedener}
\end{figure*}

@@

It is a general belief~\cite{Rudd1992} that the 
angular dispersion of emitted electrons is nearly isotropic. This 
behavior is thought to be caused by the insignificant angular anisotropy 
of sub--50--eV yield. 
A typical value for the ejection angle considered in the literature is 
$\overline{\theta}_{\alpha}\sim 70\degree$~\cite{surdutovic2018}, and 
it is quite correct in the range of validity of the first Born 
approximation for any target. However, when a distorted wave approximation 
is used, $\overline{\theta}_{\alpha}$ decreases substantially with $Z$ 
in the intermediate energy region. This effect is evident in 
Fig.~\ref{fig:emittedang};
for example, at 0.3 MeV/amu --where the Bragg peak for C$^{6+}$ impact 
occurs-- the CDW $\overline{\theta}_{\alpha}$ is half of the first Born 
value. This correction closes the damage to the forward direction.
We can attribute this forward direction correction to the capture to 
the continuum effect; the higher the charge $Z$, the smaller 
$\overline{\theta}$ will be. Of course, this effect only holds at 
intermediate energies and not at high impact energies.

@@

In Fig.~\ref{fig:crossDNA_1}, we report the total ionization cross 
sections by the impact of multicharged ions for adenine, cytosine, 
guanine, and thymine, by combining the SSM and CDW results in 
Eq.~\ref{eq:sumion}. For adenine, the agreement with the experimental 
data available for proton impact~\cite{iriki2011} is excellent. To the 
best of our knowledge, there are no experimental data on ion--collision 
ionization for the rest of the molecules. We have also included in this 
figure electron impact measurements~\cite{rahman2016} with the 
corresponding equivelocity conversion for electron incident energies 
higher than 300~eV. In this region, the proton and electron cross 
section should converge. Although the electron impact measurements are 
above our findings for all the molecular targets, it is worth stating 
that our results agree very well with other electron impact theoretical 
predictions~\cite{mozejko2003,tan2018}.

Our results for uracil, DNA backbone, pyrimidine, and THF are displayed 
in Fig.~\ref{fig:crossDNA_2}. For uracil, the agreement with the 
experimental proton impact measurements by 
Itoh~{\it et al.}~\cite{itoh2013} is good.
However, for the same target, our theory is a factor of two above 
the experimental ionization measurements by Tribedi and 
collaborators~\cite{agnihotri2012,agnihotri2013} by the impact of 
multicharged ions.
Nonetheless, it should be stated that our theoretical results coincide 
with calculations by Champion, Rivarola, and 
collaborators~\cite{agnihotri2012,champion2012}, which may indicate a 
possible misstep of the experiments.

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/adn1.eps}
\caption{Reduced CDW ionization cross section by the impact of 
multicharged ions. Experiments: \mbox{\Large$\circ$}~\cite{iriki2011} 
for proton impact and $\square$~\cite{rahman2016} for electron impact 
with equivelocity conversion.}
\label{fig:crossDNA_1}
\end{figure*}

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/adn2.eps}
\caption{Reduced CDW ionization cross section by the impact of 
multicharged ions. Experiments: proton impact on 
$\triangle$ uracil~\cite{itoh2013}, 
$\bigtriangledown$ pyrimidine~\cite{wolff2014} and $\meddiamond$
THF~\cite{wang2016}. Impact of $\ominus$ C$^{4+}$, 
$\oplus$ C$^{6+}$, O$^{6+}$, F$^{6+}$, and
$\otimes$ O$^{8+}$, F$^{8+}$ on 
uracil~\cite{agnihotri2012,agnihotri2013}. 
Symbols~$\rhd$~\cite{bug2017}, $\lhd$~\cite{wolf2019}, and 
$\medstar$~\cite{fuss2009} for electron impact with equivelocity 
conversion.}
\label{fig:crossDNA_2}
\end{figure*}

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{|p{0.07\textwidth}p{0.02\textwidth}p{0.02\textwidth}|
p{0.095\textwidth}p{0.02\textwidth}p{0.02\textwidth}|
p{0.095\textwidth}p{0.035\textwidth}p{0.02\textwidth}|}
\hline
 Molecule & $n_e'$ & $n_e$ & Molecule          & $n_e'$ & $n_e$ & Molecule             & $n_e'$ & $n_e$ \\
\hline
 H$_2$    & 2 & 2   & C$_2$H$_7$N         & 19 & 20 & C$_4$H$_5$N$_3$O     & 37 & 42 \\
 H$_2$O   & 6 & 8   & C$_4$H$_8$O         & 28 & 30 & C$_5$H$_6$N$_2$O$_2$ & 42 & 48 \\
 NH$_3$   & 7 & 8   & C$_4$H$_4$N$_2$     & 28 & 30 & C$_5$H$_5$N$_5$      & 45 & 50 \\
 CH$_4$   & 8 & 8   & C$_6$H$_6$          & 30 & 30 & C$_5$H$_5$N$_5$O     & 49 & 56 \\
 CH$_5$N  & 13 & 14 & C$_4$H$_4$N$_2$O$_2$& 36 & 40 & C$_5$H$_{10}$O$_5$P  & 54.5 & 65 \\
 \hline
\end{tabular}
\caption{New scaling numbers, $n_e'$, and Toburen numbers, $n_e$,
for some molecular targets of biological interest.}
\label{nn}
\end{center}
\end{table}

@@

By using Eq.~(\ref{eq:scalingCDW}), we define new active electron 
numbers $n_e'$ for some molecules of interest in Table~\ref{nn}. These 
values are different from the ones proposed by Toburen and used by 
other authors~\cite{itoh2013}. Moreover, an alternative way of showing 
the scaling can be attained by plotting the ionization cross sections 
of molecules as a function of the number of active electrons from 
Table~\ref{nn}. Our findings are displayed in Fig.~\ref{fig:recta} for 
impact energies 0.5, 1, and 2 MeV. As can be noted, the computed CDW 
ionization cross sections for all the molecules show a linear 
dependence with the number of electrons from Table~\ref{nn}. We obtain 
similar results, even for $E=10$~MeV. The comparison with the 
experimental data available shows overall good agreement, for the 
smallest molecules, H$_2$, H$_2$O, and CH$_4$, up to the most complex 
ones, like adenine. For electron impact data, the experimental data was 
interpolated between close neighbors.

@@

@@

>>>>

The ionization of biological molecules by multicharged ions constitutes 
the primary damage mechanism. The most widely used method to predict
such processes is the first Born approximation. At high energies, this 
perturbative method warrants the $Z^{2}$ laws, where $Z$ is the 
projectile charge. However, the damage is concentrated in the 
vicinities of the Bragg peak --at energies of hundreds of keV/amu--, 
precisely where the Born approximation starts to fail. 
Another theoretical issue arises due to the targets themselves; we are 
dealing with complex molecules, and the description of such targets 
represents a hard task for {\it ab initio} calculations.

Different approaches have been proposed to deal with the ionization of 
molecular targets within the independent atom model. For example, 
Galassi \textit{et al.} \cite{galassi2000} obtains molecular cross 
sections by combining CDW-EIS atomic ones based on the population of 
the molecular orbitals. Much more recently, L\"udde \textit{et al.} 
\cite{ludde2016,ludde2018} 
propose a combination of atomic cross sections with geometrical screening
corrections.

The objective of this article is to deal with two aspects; first, we perform 
more appropriate calculations on the primary damage mechanism, which can 
replace the Born results. Second, we inspect and test a stoichiometric 
model to describe the ionization of molecular targets.

@@

The total ionization cross sections of these atoms $\sigma_{\alpha}$
were calculated using the CDW. The initial bound and final continuum 
radial wave functions were obtained by using the {\sc radialf} code, 
developed by Salvat and co--workers~\cite{salvat1995}, and a Hartree--Fock 
potential obtained from the Depurated Inversion Method~\cite{mendez2016,mendez2018}. 
We used a few thousand pivot points to solve the Schr\"{o}dinger 
equation, depending on the number of oscillations of the continuum 
state. The radial integration was performed using the cubic spline 
technique. We expand our final continuum wave function as usual,
\begin{equation}
\psi_{\overrightarrow{k}}^{-}(\overrightarrow{r})=\sum_{l=0}^{l_{\max
}}\sum_{m=-l}^{l}R_{kl}^{-}(r)Y_{l}^{m}(\widehat{r})Y_{l}^{m^{\ast }}
(\widehat{k})\,.
\label{eq:contwave}
\end{equation}
The number of angular momenta considered varied from 8, at 
very low ejected--electron energies, up to $l_{\max}\sim 30$, 
for the highest energies 
considered. The same number of azimuth angles were required to obtain 
the four--fold differential cross section. The calculation performed does 
not display prior--post discrepancies at all. Each atomic total cross 
section was calculated using 35 to 100 momentum transfer values, 28 
fixed electron angles, and around 45 electron energies depending on the 
projectile impact energy. 
Further details of the calculation are given in Ref.~\cite{montanari2017}.

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/atomicscaling.eps}
\caption{Reduced CDW total ionization cross section $\sigma_{\alpha}/Z^2$ 
of six atomic 
targets. The curves are labeled with the charge state corresponding to 
the six multicharged projectiles.}
\label{fig:atomscaling}
\end{figure*}

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/ener_mean.eps}
\caption{Mean emitted energy distribution for ionization by the impact 
of multicharged ions, given by Eq.~(\ref{40}). 
Solid lines for ion charges $+1$, $2+$, $4+$, $6+$ and $8+$, as indicated.
Dashed lines for $\bar{p}$ and dotted line for the Born approximation with $Z=1$.}
\label{fig:emittedener}
\end{figure*}

@@

It is a general belief~\cite{Rudd1992} that the 
angular dispersion of emitted electrons is nearly isotropic. This 
behavior is thought to be caused by the insignificant angular anisotropy 
of sub--50--eV yield. 
A typical value for the ejection angle considered in the literature is 
$\overline{\theta}_{\alpha}\sim 70\degree$~\cite{surdutovic2018}, and 
it is quite correct in the range of validity of the first Born 
approximation for any target. However, when a distorted wave approximation 
is used, $\overline{\theta}_{\alpha}$ decreases substantially with $Z$ 
in the intermediate energy region, as shown in Fig.~\ref{fig:emittedang}.

To illustrate these effects, consider the impact of C$^{6+}$ impact on oxygen.
At an impact energy of 500~keV, the first Born approximation predicts
emitted electrons with mean energies of 46.7 eV and mean angles of 78\degree, 
while the CDW gives 62.5 eV and 60\degree. 
These results imply depper penetration closer to the direction of the ion.
We can attribute this forward direction correction to the capture to 
the continuum effect; the higher the charge $Z$, the smaller 
$\overline{\theta}$ will be. Of course, this effect only holds at 
intermediate energies and not at high impact energies.

@@

In Fig.~\ref{fig:crossDNA_1}, we report the reduced total ionization cross 
sections $\sigma_M/Z^2$ by the impact of multicharged ions for adenine,
cytosine, 
guanine, and thymine, by combining the SSM and CDW results in 
Eq.~\ref{eq:sumion}. For adenine, the agreement with the experimental 
data available for proton impact~\cite{iriki2011} is excellent. To the 
best of our knowledge, there are no experimental data on ion--collision 
ionization for the rest of the molecules. We have also included in this 
figure electron impact measurements~\cite{rahman2016} with the 
corresponding equivelocity conversion for electron incident energies 
higher than 300~eV. In this region, the proton and electron cross 
section should converge. Although the electron impact measurements are 
above our findings for all the molecular targets, it is worth stating 
that our results agree very well with other electron impact theoretical 
predictions~\cite{mozejko2003,tan2018}.

At intermediate impact energies, the $Z^2$ rule no longer holds and other 
scalings can be considered in this region. 
In Fig.~\ref{fig:Zcaled_crossDNA_1}, we plotted the reduced ionization 
cross section $\sigma_M/Z$ as a function of the reduced ion impact energy
$E/Z$ as suggested by~\cite{janev1980,dubois2013}.
This scaling may be a good alternative to deal with highly charged
ions at intermediate energies in a coherent way.

The reduced total ionization cross sections $\sigma_M/Z^2$ for uracil, 
DNA backbone, pyrimidine, and THF are displayed 
in Fig.~\ref{fig:crossDNA_2}. For uracil, the agreement with the 
experimental proton impact measurements by 
Itoh~{\it et al.}~\cite{itoh2013} is good.
However, for the same target, our theory is a factor of two above 
the experimental ionization measurements by Tribedi and 
collaborators~\cite{agnihotri2012,agnihotri2013} by the impact of 
multicharged ions.
Nonetheless, it should be stated that our theoretical results coincide 
with calculations by Champion, Rivarola, and 
collaborators~\cite{agnihotri2012,champion2012}, which may indicate a 
possible misstep of the experiments.

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/adn1.eps}
\caption{Reduced CDW ionization cross section $\sigma_{M}/Z^2$ as a function 
of ion impact energy. Experiments: \mbox{\Large$\circ$}~\cite{iriki2011} 
for proton impact and $\square$~\cite{rahman2016} for electron impact 
with equivelocity conversion.}
\label{fig:crossDNA_1}
\end{figure*}

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/Zscaling_adn1.eps}
\caption{Reduced CDW ionization cross section $\sigma_{M}/Z$ as a function 
of reduced ion impact energy $E/Z$. 
Experiments: \mbox{\Large$\circ$}~\cite{iriki2011} 
for proton impact and $\square$~\cite{rahman2016} for electron impact 
with equivelocity conversion.}
\label{fig:Zcaled_crossDNA_1}
\end{figure*}

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/adn2.eps}
\caption{Reduced CDW ionization cross section $\sigma_{M}/Z^2$ as a function 
of ion impact energy. Experiments: proton impact on 
$\triangle$ uracil~\cite{itoh2013}, 
$\bigtriangledown$ pyrimidine~\cite{wolff2014} and $\meddiamond$
THF~\cite{wang2016}. Impact of $\ominus$ C$^{4+}$, 
$\oplus$ C$^{6+}$, O$^{6+}$, F$^{6+}$, and
$\otimes$ O$^{8+}$, F$^{8+}$ on 
uracil~\cite{agnihotri2012,agnihotri2013}. 
Symbols~$\rhd$~\cite{bug2017}, $\lhd$~\cite{wolf2019}, and 
$\medstar$~\cite{fuss2009} for electron impact with equivelocity 
conversion.}
\label{fig:crossDNA_2}
\end{figure*}

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{|p{0.07\textwidth}p{0.02\textwidth}p{0.02\textwidth}|
p{0.095\textwidth}p{0.02\textwidth}p{0.02\textwidth}|
p{0.095\textwidth}p{0.035\textwidth}p{0.02\textwidth}|}
\hline
 Molecule & $n_e'$ & $n_e$ & Molecule          & $n_e'$ & $n_e$ & Molecule             & $n_e'$ & $n_e$ \\
\hline
 H$_2$    & 2 & 2   & C$_2$H$_7$N         & 19 & 20 & C$_4$H$_5$N$_3$O     & 37 & 42 \\
 H$_2$O   & 6 & 8   & C$_4$H$_8$O         & 28 & 30 & C$_5$H$_6$N$_2$O$_2$ & 42 & 48 \\
 NH$_3$   & 7 & 8   & C$_4$H$_4$N$_2$     & 28 & 30 & C$_5$H$_5$N$_5$      & 45 & 50 \\
 CH$_4$   & 8 & 8   & C$_6$H$_6$          & 30 & 30 & C$_5$H$_5$N$_5$O     & 49 & 56 \\
 CH$_5$N  & 13 & 14 & C$_4$H$_4$N$_2$O$_2$& 36 & 40 & C$_5$H$_{10}$O$_5$P  & 54.5 & 65 \\
 \hline
\end{tabular}
\caption{New scaling numbers $n_e'$, and Toburen numbers $n_e$,
for some molecular targets of biological interest.}
\label{nn}
\end{center}
\end{table}

@@

By using Eq.~(\ref{eq:scalingCDW}), we define new active electron 
numbers $n_e'$ for some molecules of interest in Table~\ref{nn}. These 
values are different from the ones proposed by Toburen and used by 
other authors~\cite{itoh2013}. Moreover, an alternative way of showing 
the scaling can be attained by plotting the ionization cross sections 
of molecules as a function of the number of active electrons from 
Table~\ref{nn}. Our findings are displayed in Fig.~\ref{fig:recta} for 
impact energies 0.5, 1, and 2 MeV. As can be noted, the computed CDW 
ionization cross sections for all the molecules show a linear 
dependence with the number of electrons $n_e'$ from Table~\ref{nn}.
We obtain similar results, even for $E=10$~MeV. An 
equivalent plot using the Toburen numbers $n_e$ of Table~\ref{nn} does not 
exhibit the straight lines obtained with the present scaling. The 
comparison with the 
experimental data available shows overall good agreement, for the 
smallest molecules, H$_2$, H$_2$O, and CH$_4$, up to the most complex 
ones, like adenine. For electron impact data, the experimental data was 
interpolated between close neighbors.

@@

\bibitem{galassi2000}
M. E. Galasssi, R. D. Rivarola, M. Beuve, G. H. Olivera and P. D. Fainstein, 
Phys. Rev. A \textbf{62}, 022701 (2000).

\bibitem{ludde2016}
H. J. L\"udde, A. Achenbach, T. Kalkbrenner, H.-C. Jankowiak and T. Kirchner,
Eur. Phys. J. D \textbf{70}, 82 (2016).

\bibitem{ludde2018}
H. J. L\"udde, M. Horbatsch and T. Kirchner,
Eur. Phys. J. B \textbf{91}, 99 (2018).

@@

\bibitem{janev1980}
R. K. Janev and L. P. Presnyakov 
J. Phys. B \textbf{13}, 4233 (1980).

\bibitem{dubois2013}
R. D. DuBois, E. C. Montenegro and G. M. Sigaud,
AIP Conference Proceeding \textbf{1525}, 679 (2013).
